Perturbation theory in the complex plane: Exceptional points and where to find them

### 2020-11-17 Manuscript/EPAWTFT.tex
Modified exact example to Wannier basis.

\documentclass[aps,prb,reprint,noshowkeys,superscriptaddress]{revtex4-1}
\usepackage{graphicx,tabularx,array,booktabs,longtable,dcolumn,xcolor,microtype,multirow,amscd,amsmath,amssymb,amsfonts,physics,siunitx,mhchem}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{txfonts}

\usepackage[normalem]{ulem}
\definecolor{darkgreen}{RGB}{0, 180, 0}
\newcommand{\titou}[1]{\textcolor{red}{#1}}
\newcommand{\hugh}[1]{\textcolor{orange}{#1}}
\newcommand{\trash}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\trashHB}[1]{\textcolor{orange}{\sout{#1}}}

\usepackage[
	colorlinks=true,
    citecolor=blue,
    breaklinks=true
	]{hyperref}
\urlstyle{same}

\newcommand{\ctab}{\multicolumn{1}{c}{---}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\mc}{\multicolumn}
\newcommand{\fnm}{\footnotemark}
\newcommand{\fnt}{\footnotetext}
\newcommand{\mcc}[1]{\multicolumn{1}{c}{#1}}
\newcommand{\mr}{\multirow}

@@

@@

%===================================%
\subsection{An illustrative example}
%===================================%
In order to highlight the general properties of EPs mentioned above, we propose to consider the ubiquitous symmetric Hubbard dimer at half filling (\ie, with two opposite-spin fermions) whose Hamiltonian reads in the singlet configuration state function basis 
%\begin{align}
%		\ket{1\up1\dw}		&	\ket{1\up2\dw}		&	\ket{1\dw2\up}		&	\ket{2\up2\dw}	\\
%		\uddot \quad \vac	&	\updot \quad \dwdot	&	\dwdot \quad \updot	&	\vac \quad \uddot	 \\
%\end{align}
\begin{equation}
\label{eq:H_FCI}
	\bH = 
	\begin{pmatrix}
		-2t	+ U &	0 & U/2	\\
		0		&	U 	&	0	\\
		U/2 &	0 & -2t	+ U	\\
	\end{pmatrix},
\end{equation}
where $t$ is the hopping parameter and $U$ is the on-site Coulomb repulsion.
We refer the interested reader to Refs.~\onlinecite{Carrascal_2015,Carrascal_2018} for more details about this system.
We will consistently use this system to illustrate the different concepts discussed in the present review article.

For real $U$, the Hamiltonian \eqref{eq:H_FCI} is clearly Hermitian, and it becomes non-Hermitian for any complex $U$ value.
The eigenvalues associated with its singlet ground state and singlet doubly-excited state are
\begin{equation}
\label{eq:E_FCI}
	E_{\pm} = \frac{1}{2} \qty( U \pm \sqrt{(4t^2) + U^2} ).
\end{equation}
and they are represented as a function of $U$ in Fig.~\ref{fig:FCI} together with the energy of the singlet open-shell configuration of energy $U$.

%%% FIG 1 %%%
\begin{figure*}
	\includegraphics[height=0.3\textwidth]{fig1a}
	\hspace{0.1\textwidth}
	\includegraphics[height=0.3\textwidth]{fig1b}	
	\caption{
	Exact energies for the Hubbard dimer as functions of $U/t$.
	\label{fig:FCI}}
\end{figure*}

One notices that these two states become degenerate only for a pair of complex conjugate values of $U$
\begin{equation}
\label{eq:lambda_EP}
	U_\text{EP} = \pm 4 i t,
\end{equation}
with energy
\begin{equation}
\label{eq:E_EP}
	E_\text{EP} = \pm 2 i t,
\end{equation}
which correspond to square-root singularities in the complex-$U$ plane [see Fig.~\eqref{fig:FCI}].
These two $U$ values, given by Eq.~\eqref{eq:lambda_EP}, are the so-called EPs and one can clearly see that they connect the ground and excited states.
Starting from $U_\text{EP}$, two square-root branch cuts run on the imaginary axis towards $\pm i \infty$.
In the real $U$ axis, the point for which the states are the closest ($U = 0$) is called an avoided crossing and this occurs at $U = \Re(U_\text{EP})$.
The ``shape'' of the avoided crossing is linked to the magnitude of $\Im(U_\text{EP})$: the smaller $\Im(U_\text{EP})$, the sharper the avoided crossing is.

Around $U = U_\text{EP}$, Eq.~\eqref{eq:E_FCI} behaves as \cite{MoiseyevBook}
\begin{equation} %\label{eq:E_EP}
        E_{\pm} = E_\text{EP} \pm \sqrt{2U_\text{EP}} \sqrt{U - U_\text{EP}},
\end{equation}
and following a complex contour around the EP, \ie, $U = U_\text{EP} + R \exp(i\theta)$, yields
\begin{equation}
        E_{\pm}(\theta) = E_\text{EP} \pm \sqrt{2U_\text{EP} R}  \exp(i\theta/2),
\end{equation}
and we have
\begin{align}
	E_{\pm}(2\pi) & = E_{\mp}(0),
	&
	E_{\pm}(4\pi) & = E_{\pm}(0). \notag
\end{align}
This evidences that encircling non-Hermitian degeneracies at EPs leads to an interconversion of states, and two loops around the EP are necessary to recover the initial energy.
Additionally, the wave function picks up a geometric phase and four loops are required to recover the starting wave function.
\cite{MoiseyevBook}

>>>>

\documentclass[aps,prb,reprint,noshowkeys,superscriptaddress]{revtex4-1}
\usepackage{subcaption}
\usepackage{graphicx,tabularx,array,booktabs,dcolumn,xcolor,microtype,multirow,amscd,amsmath,amssymb,amsfonts,physics,siunitx,mhchem}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{txfonts}

\usepackage[normalem]{ulem}
\definecolor{hughgreen}{RGB}{0, 128, 0}
\newcommand{\titou}[1]{\textcolor{red}{#1}}
\newcommand{\hugh}[1]{\textcolor{hughgreen}{#1}}
\newcommand{\trash}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\trashHB}[1]{\textcolor{orange}{\sout{#1}}}

\usepackage[
	colorlinks=true,
    citecolor=blue,
    linkcolor=blue,
    filecolor=blue,      
    urlcolor=blue,	
    breaklinks=true
	]{hyperref}
\urlstyle{same}

\newcommand{\ctab}{\multicolumn{1}{c}{---}}
\newcommand{\latin}[1]{#1}
%\newcommand{\latin}[1]{\textit{#1}}
\newcommand{\ie}{\latin{i.e.}}
\newcommand{\eg}{\latin{e.g.}}
\newcommand{\mc}{\multicolumn}
\newcommand{\fnm}{\footnotemark}
\newcommand{\fnt}{\footnotetext}
\newcommand{\mcc}[1]{\multicolumn{1}{c}{#1}}
\newcommand{\mr}{\multirow}

@@

% Center tabularx columns
\newcolumntype{Y}{>{\centering\arraybackslash}X}

% Imaginary constant
\renewcommand{\i}{\mathrm{i}}
% Blackboard bold
\newcommand{\bbR}{\mathbb{R}}
\newcommand{\bbC}{\mathbb{C}}

\newcommand{\Lup}{\text{L}^{\uparrow}}
\newcommand{\Ldown}{\text{L}^{\downarrow}}
\newcommand{\Rup}{\text{R}^{\uparrow}}
\newcommand{\Rdown}{\text{R}^{\downarrow}}

@@

%===================================%
\subsection{\hugh{Exact Exceptional Points}}
%===================================%

%%% FIG 1 %%%
\begin{figure*}[t]
	\begin{subfigure}{0.49\textwidth}
	\includegraphics[height=0.65\textwidth]{fig1a}
	\subcaption{\label{subfig:FCI_real}}
    \end{subfigure}
	\begin{subfigure}{0.49\textwidth}
	\includegraphics[height=0.65\textwidth]{fig1b}
	\subcaption{\label{subfig:FCI_cplx}}
    \end{subfigure}
	\caption{%
	\hugh{Exact energies for the Hubbard dimer ($U=4t$) as functions of $\lambda$ on the real axis (\subref{subfig:FCI_real}) and in the complex plane (\subref{subfig:FCI_cplx}).
    Only the interacting closed-shell singlets are plotted in the complex plane, becoming degenerate at the EP (black dot).}
	\label{fig:FCI}}
\end{figure*}

\hugh{To illustrate the concepts discussed throughout this article, we will consider the symmetric Hubbard dimer at half filling, \ie\ with two opposite-spin fermions.
Using the localised Wannier basis, the Hilbert space for this system comprises the four configurations
\begin{equation}
\ket{\Lup \Ldown} \qquad \ket{\Lup\Rdown} \qquad \ket{\Rup\Ldown} \qquad \ket{\Rup\Rdown}
\end{equation}
%\begin{tabularx}{\linewidth}{YYYY}
%$\ket{\Lup \Ldown}$ & $\ket{\Lup\Rdown}$ & $\ket{\Rup\Ldown}$ &	$\ket{\Rup\Rdown}$
%\\
%$\uddot \quad \vac$ & $\updot \quad \dwdot$ & $\dwdot \quad \updot$ & $\vac \quad \uddot$
%\end{tabularx}
where $\text{L}^{\sigma}$ ($\text{R}^{\sigma}$) denotes an electron with spin $\sigma$ on the left (right) site.
The exact Hamiltonian is then 
\begin{equation}
\label{eq:H_FCI}
	\bH = 
	\begin{pmatrix}
		U &	- t & -  t & 0	\\
	   -  t &  0 &  0 & -  t \\
       -  t &  0 &  0 & -  t \\
        0 & -  t & -  t & U \\
	\end{pmatrix},
\end{equation}
where $t$ is the hopping parameter and $U$ is the on-site Coulomb repulsion.}
We refer the interested reader to Refs.~\onlinecite{Carrascal_2015,Carrascal_2018} for more details about this system.

\hugh{%
To illustrate the formation of an exceptional points, we scale the off-diagonal coupling strength by introducing the complex parameter $\lambda$ through the transformation $t\rightarrow \lambda t$.
When $\lambda$ is real, the Hamiltonian~\eqref{eq:H_FCI} is Hermitian with the distinct (real-valued) eigenvalues
\begin{subequations}
\begin{align}
E_{\mp} &= \frac{1}{2} \qty(U \mp \sqrt{ (4 \lambda t)^2 + U^2 } ),
\label{eq:singletE}
\\
E_{\text{T}} &= 0,
\\
E_{\text{S}} &= U.
\end{align}
\end{subequations}
While the open-shell triplet ($E_{\text{T}}$) and singlet ($E_{\text{S}}$) are independent of $\lambda$, the closed-shell singlet ground state ($E_{-}$) and doubly-excited state ($E_{+}$) couple strongly to form an avoided crossing at $\lambda=0$ (see Figure~\ref{subfig:FCI_real}).
At non-zero values of $U$ and $t$, these closed-shell singlets can only become degenerate at a pair of complex conjugate points in the complex $\lambda$ plane 
\begin{equation}
\lambda_{\text{EP}} = \pm \frac{U}{4t} \i,
\end{equation}
with energy
\begin{equation}
\label{eq:E_EP}
	E_\text{EP} = \frac{U}{2}.
\end{equation}
These $\lambda$ values correspond to so-called EPs and connect the ground and excited state in the complex plane.
Crucially, the energy surface becomes non-analytic at $\lambda_{\text{EP}}$ and a square-root singularity forms with two branch cuts running along the imaginary axis from $\lambda_{\text{EP}}$  to $\pm \i \infty$ (see Fig.~\ref{subfig:FCI_cplx}).
On the real $\lambda$ axis, these EPs lead to the singlet avoided crossing at $\lambda = \Re(\lambda_{\text{EP}})$.
The ``shape'' of this avoided crossing is related to the magnitude of $\Im(\lambda_{\text{EP}})$, with smaller values giving a ``sharper'' interaction.
}

\hugh{
Remarkably, the existence of these square-root singularities means that following a complex contour around an EP in the complex $\lambda$ plane will interconvert the closed-shell ground and excited states.
This behaviour can be seen by expanding the radicand in Eq.~\eqref{eq:singletE} as a Taylor series around $\lambda_{\text{EP}}$ to give
\begin{equation}
E_{\pm} \approx E_{\text{EP}} \pm \sqrt{32t^2 \lambda_{\text{EP}}} \sqrt{\lambda - \lambda_{\text{EP}}}.
\end{equation}
Parametrising the complex contour as $\lambda(\theta) = \lambda_{\text{EP}} + R \exp(\i \theta)$ gives the continuous energy pathways 
\begin{equation}
E_{\pm} \qty(\theta) \approx E_{\text{EP}} \pm \sqrt{32t^2 \lambda_{\text{EP}} R}\, \exp(\i \theta/2)
\end{equation}
such that}
\begin{align}
	E_{\pm}(2\pi) & = E_{\mp}(0),
	&
	E_{\pm}(4\pi) & = E_{\pm}(0). \notag
\end{align}
As a result, completely encircling an EP leads to the interconversion of the two interacting states, while a second complete rotation returns the two states to their original energies.
Additionally, the wave functions pick up a geometric phase in the process, and four complete loops are required to recover their starting forms.\cite{MoiseyevBook}

%In order to highlight the general properties of EPs mentioned above, we propose to consider the ubiquitous symmetric Hubbard dimer at half filling (\ie, with two opposite-spin fermions) whose Hamiltonian reads in the singlet configuration state function basis 
%\begin{align}
%		\ket{1\up1\dw}		&	\ket{1\up2\dw}		&	\ket{1\dw2\up}		&	\ket{2\up2\dw}	\\
%		\uddot \quad \vac	&	\updot \quad \dwdot	&	\dwdot \quad \updot	&	\vac \quad \uddot	 \\
%\end{align}
%\begin{equation}
%\label{eq:H_FCI}
%	\bH = 
%	\begin{pmatrix}
%		-2t	+ U &	0 & U/2	\\
%		0		&	U 	&	0	\\
%		U/2 &	0 & -2t	+ U	\\
%	\end{pmatrix},
%\end{equation}
%where $t$ is the hopping parameter and $U$ is the on-site Coulomb repulsion.
%We refer the interested reader to Refs.~\onlinecite{Carrascal_2015,Carrascal_2018} for more details about this system.
%We will consistently use this system to illustrate the different concepts discussed in the present review article.

%For real $U$, the Hamiltonian \eqref{eq:H_FCI} is clearly Hermitian, and it becomes non-Hermitian %for any complex $U$ value.
%The eigenvalues associated with its singlet ground state and singlet doubly-excited state are
%\begin{equation}
%\label{eq:E_FCI}
%	E_{\pm} = \frac{1}{2} \qty( U \pm \sqrt{(4t^2) + U^2} ).
%\end{equation}
%and they are represented as a function of $U$ in Fig.~\ref{fig:FCI} together with the energy of the %singlet open-shell configuration of energy $U$.

%One notices that these two states become degenerate only for a pair of complex conjugate values of $U$
%\begin{equation}
%\label{eq:lambda_EP}
%	U_\text{EP} = \pm 4 i t,
%\end{equation}
%with energy
%\begin{equation}
%\label{eq:E_EP}
%	E_\text{EP} = \pm 2 i t,
%\end{equation}
%which correspond to square-root singularities in the complex-$U$ plane [see Fig.~\eqref{fig:FCI}].
%These two $U$ values, given by Eq.~\eqref{eq:lambda_EP}, are the so-called EPs and one can clearly see that they connect the ground and excited states.
%Starting from $U_\text{EP}$, two square-root branch cuts run on the imaginary axis towards $\pm i \infty$.
%In the real $U$ axis, the point for which the states are the closest ($U = 0$) is called an avoided crossing and this occurs at $U = \Re(U_\text{EP})$.
%The ``shape'' of the avoided crossing is linked to the magnitude of $\Im(U_\text{EP})$: the smaller $\Im(U_\text{EP})$, the sharper the avoided crossing is.

%Around $U = U_\text{EP}$, Eq.~\eqref{eq:E_FCI} behaves as \cite{MoiseyevBook}
%\begin{equation} %\label{eq:E_EP}
%        E_{\pm} = E_\text{EP} \pm \sqrt{2U_\text{EP}} \sqrt{U - U_\text{EP}},
%\end{equation}
%and following a complex contour around the EP, \ie, $U = U_\text{EP} + R \exp(i\theta)$, yields
%\begin{equation}
%        E_{\pm}(\theta) = E_\text{EP} \pm \sqrt{2U_\text{EP} R}  \exp(i\theta/2),
%\end{equation}
%and we have
%\begin{align}
%	E_{\pm}(2\pi) & = E_{\mp}(0),
%	&
%	E_{\pm}(4\pi) & = E_{\pm}(0). \notag
%\end{align}
%This evidences that encircling non-Hermitian degeneracies at EPs leads to an interconversion of states, and two loops around the EP are necessary to recover the initial energy.
%Additionally, the wave function picks up a geometric phase and four loops are required to recover the starting wave function.
%\cite{MoiseyevBook}
